---
title: "Lecture - Week 4"
author: "Ziq"
date: "March 24, 2019"
output: html_document
---

```{r}
#basically in this dataset, we want to separate the activities based on the features that they were performing.
load('samsungData.rda')
names(samsungData)[1:12]
table(samsungData$activity)
```

So first thing we can do is plot average accelleration for the first subject.

```{r}
par(mfrow = c(1,2), mar = c(5,4,1,1))
samsungData <- transform(samsungData, activity = factor(activity))
sub1 <- subset(samsungData, subject == 1)
plot(sub1[, 1], col = sub1$activity, ylab = names(sub1)[1]) #body acceleration in the mean X
plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[1]) #body acceleration in the mean Y
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), pch = 1)

distanceMatrix <- dist(sub1[,1:3])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
```

So what if we looked at max acceleration?

```{r}
par(mfrow = c(1,2))
plot(sub1[, 10], pch = 19, col = sub1$activity, ylab = names(sub1)[10]) #max accel x
plot(sub1[, 11], pch = 19, col = sub1$activity, ylab = names(sub1)[11]) #max accel y
distanceMatrix <- dist(sub1[, 10:12])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity)) #now we can see that clusters of moving actions are closer together
```
## Singular Value Decomposition

```{r}
svd1 = svd(scale(sub1[, -c(562, 563)]))
par(mfrow = c(1, 2))
plot(svd1$u[,1], col = sub1$activity, pch = 19) #really separates moving from non-moving
plot(svd1$u[,2], col = sub1$activity, pch = 19) #kinda vague, magenta seems to be separated from everything? 
#let's try to figure out what the biggest contributor in the variation for features for the different observations
maxContrib <- which.max(svd1$v[,2])
distanceMatrix <- dist(sub1[, c(10:12, maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity)) #didn't really help to spearate the non-moving activities
#but it helped a lot for separating the moving activities.
names(samsungData)[maxContrib]
```

## K-Means clustering

nstart = 1, first try
nstart is number of starting values and taking the most optimal ones.

```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6) #remove activity and subject cuz we don't need it, start random centers
table(kClust$cluster, sub1$activity) #clusters are the rows. K-means had a bit of trouble separating laying sitting and standing.
kClust <- kmeans(sub1[, -c(562, 563)], centers = 100)
```

So we can see where the clusters are. We can see which features drive the center of the cluster, what features are important for classifying activities in that custer?

```{r}
plot(kClust$center[4, 1:10], pch = 19, ylab = "Cluster Center", xlab = "") #for walking, we see that these features
# are more important than others.
```


## Helper function

```{r}
myplclust <- function( hclust, lab=hclust$labels, lab.col=rep(1,length(hclust$labels)), hang=0.1,...){
 ## modifiction of plclust for plotting hclust objects *in colour*!
 ## Copyright Eva KF Chan 2009
 ## Arguments:
 ##    hclust:    hclust object
 ##    lab:        a character vector of labels of the leaves of the tree
 ##    lab.col:    colour for the labels; NA=default device foreground colour
 ##    hang:     as in hclust & plclust
 ## Side effect:
 ##    A display of hierarchical cluster with coloured leaf labels.
 y <- rep(hclust$height,2)
 x <- as.numeric(hclust$merge)
 y <- y[which(x<0)]
 x <- x[which(x<0)]
 x <- abs(x)
 y <- y[order(x)]
 x <- x[order(x)]
 plot( hclust, labels=FALSE, hang=hang, ... )
 text( x=x, y=y[hclust$order]-(max(hclust$height)*hang), labels=lab[hclust$order], col=lab.col[hclust$order], srt=90, adj=c(1,0.5), xpd=NA, ... )}

```

