---
title: "R Cheatbook (Getting and Cleaning Data)"
output: rmarkdown::github_document
---

This is an R notebook for me to reference to when I decide to clean or sort data. However, 
after learning the DPLYR package, I'm not so confident I will use most of this.

## Subsetting and Sorting

data.frame() column variables must be the same atomic class
```{r}
df <- data.frame("var1" = c(2,1,3), "var20" = c('b', 'a', 'c'))
df
```

To subset, we use the square brakets [] after a dataframe e.g. df[row, column]. Square brackets can take index values as well as logical values including ands and ors. 
```{r}
df[df$var1 > 2,]
```
However, if we're dealing with NA values, then we'd want to use the which function which returns positions and indices.
```{r}
df[which(df$var1 > 2),]
```
We can also order rows using the order funcion.

```{r}
df[order(df$var1),]
```

## Summarizing Data

Summary Sample output

```{r}
library(datasets)
data(iris)
summary(iris)
```
c
Summary is pretty good for informing about NA values.

##Cross Tabs

```{r}
data(UCBAdmissions)
df <- as.data.frame(UCBAdmissions)
head(df, 5) 
xt <- xtabs(Freq ~ Gender + Admit, data=df)
xt
```
## New Variables
df$newvar <- c(...)
df$newvar <- ifelse(x < y, TRUE, FALSE)


