---
title: "Lecture - Week 2"
author: "Ziq"
date: "March 29, 2019"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

## Coding Standards

Column margins at 80, tab length at least 4, use an IDE and make your functions 
short.

## Markdown

It's a text-to-HTML conversion tool for web writers. 

#### Syntax

*italicize*
**bold**

- first item in list
- second item in list
- third item in list

[link to google](google.ca)

[link to google] [1] and [link to hotmail] [2]

A spacing is needed before the quotes.

[1]: https://www.google.ca/ "google"
[2]: https://www.hotmail.com/ "hotmail"

Double space after a sentence will make a new line  
assfas

### What is Markdown

## Knitr

So authors had taken considerable effort to put data/results onto the web. Knitr
sort of solves this problem by providing this basic idea where an article is a 
stream of text and code.

It's for a person to focus on writing as opposed to formatting. It'd be also
be really useful to do this from the start. The cons ot literate programing is 
the text and code are in one place so it's a lengthy analysis.

### What is knitr?

Already built into R studio which is nice. Markdown is a simplified version
of markup languages. It's good for manuals, short or medium technical 
documents, tutorials, repors and data preprocessing documents/summaries.

Not great for long research articles, time consuming computations, or
documents that require precies formatting.

```{r}
1 + 1
```

If you just want the output then
```{r, echo=FALSE}
1 + 1
```

if you just want the code then

```{r, results='hide'}
1 + 1
```

```{r ,echo = FALSE}
time <- format(Sys.time(), "%a %b %d %X %Y")
string <- "cheese"
```

The current time is `r time`. I like `r string`

```{r}
opts_chunk$set(echo = false, results = "hide")
```

One chunk can somtimes take al ong time to run and chunks have to be recomputed
everytime you re-knit a file. Setting cache=TRUE can be set to store results of
computation.


