---
title: "DPLYR"
output: rmarkdown::github_document
---

All DPLYR functions require that that you enter a data frame as its first argument, and will always return a data frame

Key Verbs:
Select: return a subset of the columns of the data
Filter: extracts a subset of rows from a data frame based on logical conditions
Arrange: reorder rows of a data frame
Rename: rename variables in a data frame
Mutate: add new variables/columns or transform existing variables
Summarise: generate summary statistics of different variable in data frame

```{r}
library(dplyr)
library(datasets)
data(iris)
str(iris)
```

#Subsetting
A lot easier because we don't need to find the indices that are required for R's basic functions.
```{r}
head(select(iris, Sepal.Length:Sepal.Width)) #or
head(select(iris, -(Sepal.Length:Sepal.Width)))
```
Filter works really similarly and really intuitively from the description.

```{r}
head(filter(iris, Sepal.Width > 3 & Sepal.Length > 5))
```

# Arrange

```{r}
head(arrange(iris, desc(Petal.Width)))
```

# Rename

```{r}
rename(iris, Sep.Length = Sepal.Length) #new name = old name
```

# Mutate

```{r}
head(mutate(iris, Normalized.SP = Sepal.Length - mean(Sepal.Length, na.rm = TRUE)))
iris <- mutate(iris, newfactor = factor(Sepal.Length > 5, labels = c("short", "long")))
lenf <- group_by(iris, newfactor)
```

#summarize
A powerful way to call a function to a dataframe that is already grouped.

```{r}
head(summarize(lenf, Sepal.Length = mean(Sepal.Length)))
```

