---
title: "Tidyr Basics"
output: rmarkdown::github_document
---

Remember that for dataframes, each row is an observation and each column is a variable (not value).
Examples are taken from the swirl package.

#Gather

Gather takes multiple columns and collapses into key-value pairs, duplicating all other columns as needed. You use gather() when you notice that you have columns that are not variables.

```{r}
students
gather(students, sex, count, male:female, -grade)
```

# Separate
separate() turns a single character column into multiple columns.  specify the first three arguments: data = res, col = sex_class, into = c("sex", "class"). separate() was able to figure out on its own how to separate the sex_class column. Unless you request otherwise with the 'sep' argument, it splits on non-alphanumeric values. In other words, it assumes that the values are separated by something other than a letter or number (in this case, an underscore.)

```{r}
students2
t1 <- gather(students2, sex_class, count, -grade)
t1
separate(t1, sex_class, c("sex", "class"))
```

More examples!
# Spread
Spread a key-value pair across multiple columns

```{r}
students3
s3.1 <- gather(students3, class, grade, class1:class5, na.rm = TRUE)
s3.1
s3.2 <- spread(s3.1, test, grade)
s3.2
mutate(s3.2, class = parse_number(class))
```

# mergedatasets
```{r}
psd <- mutate(passed, status = "passed")
fld <- mutate(failed, status = "failed")
psd
fld
bind_rows(psd,fld)
```

